Distance Sampling in \proglang{R}

### 2016-01-27 paper.Rmd
Further editing on Intro section 1
Now I have edited the first part of the intro I am not sure if Section
1.1 wants to go in a different direction rather than being a basic
intro to distance sampling. I think that the point transect figures
were a bit confusing without quite a bit more explanation and I’m not
sure we want to go there but could still talk about line transects. But
I did wonder if section 1.1 wanted to focus on why use our distance
package rather than the other options? What do you all think (do you
like the intro after the restructuring and if so what should section
1.1 focus o

(LT: I wonder who the intended audience is for this paper? (LM: - Distance for Windows users and other people looking do distance sampling in R?) The way it is currently written, it ``kinda'' introduces distance sampling as well as the relevant commands.  Is this standard for a paper in this journal -- to introduce the methods as well as the software?  Can we assume a reasonable level of analytical sophistication on the part of the reader, and so start with a few more formulae?  Can we assume they know the methods, more or less, or point them at somewher where this is already introduced.  (Perhaps I should go back and have a look at the EcolApps paper to see what approach we took there... (LM: The paper on Distance in JAppEcol? There is quite a big intro on distance sampling there.)))

Distance sampling [@Buckland:2001vm; @Buckland:2004ts; @buckland2015distance] encompasses a suite of methods used to estimate the density and/or abundance of biological populations. Distance sampling can be thought of as an extension of plot sampling. Plot sampling involves selecting a number of plots (small areas) at random within the study area and counting the objects of interest that are contained within each plot. By selecting the plots at random we can assume that the density of objects in the plots is representative of the study area as a whole. One of the key assumptions of plot sampling is that all objects within each of the plots are counted. Distance sampling relaxes this assumption in that observers are no longer required to see and count everything within selected plots.

Conventional distance sampling assumes the observer is either stood at a point or moving along a line and will only see everything that occurs exactly at that point or on the line. It is then expected that the further away an object is from the point or line (the sampler) the less likely it is that the observer will see it. By recording the distance to each of the detected objects we can fit a model to describe the probability of detection given distance from the sampler, which we refer to as the ‘detection function’.  The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance.

The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") can be used to fit detection functions to such distance sampling data. It was first released in November 1998 as Distance for Windows version 3.5 and was initially based on its predecessor “TRANSECT”. Since this time it has evolved to include various design and analysis features. Distance for Windows versions 5 onwards have included \proglang{R} [@rcore]  packages as the analysis engines providing additional more complex analysis options than those offered by the original FROTRAN engine. Further details of the history of Distance for Windows are included in @Thomas:2010cf.

As Distance for Windows becomes increasingly reliant on analyses performed in \proglang{R} and new methods are being developed at a rate where it is unfeasible to make all these available in the Windows interface, we are encouraging the use of our \proglang{R} packages directly. Until now those wishing to use \proglang{R} for straight forward distance sampling analyses would have had to negotiate the complex package \pkg{mrds} [@mrds-pkg] designed for mark-recapture distance sampling. \pkg{Distance} is a wrapper package around \pkg{mrds} making it easier to get started with basic distance sampling analyses in \proglang{R}. We demonstrate how to use \pkg{Distance} to fit detection functions, perform model checking and selection, and estimate abundance and/or density.

## Distance sampling

(LT: a census is a complete count; I don't really think that sample surveys fall under this.  So, a bit of editing needed here.) Census-type surveys (quadrat or strip transects assuming perfect detectability within strips or quadrats) are inefficient (requiring considerable field effort) and we should expect that not all objects (animals, plants, dung, etc) can be observed. Accounting for imperfect detectability is an important consideration when obtaining accurate estimates of abundance [@LahozMonfort:2013bg]. Using the extra information gained by recording distance from the sampler to the observation, it is possible to model detectability. Because we expect detectability to decrease with increasing distance from the sampler, we model detectability as a function of distance (plus perhaps other covariates, see below). (LT: The word detection function is used again without it having been defined explicitly.) We use the probabilities of detection calculated from the detection function to estimate abundance in the area covered by the survey, which we then scale-up to the study region of interest.

Distance sampling surveys are conducted using two transect types: line and point transects. In line transect sampling observers walk (or fly, sail, etc)  lines, observing objects and recording their perpendicular distance to the line; whereas in point transect sampling observers remain stationary at a location and record distances to objects from that point. Field methods are chosen to be suitable to species and habitat constraints [@buckland2015distance].

```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects; blue lines (top plot) and crosses (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals. The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\\label{fig:pointslines}"}
opar <- par(mfrow=c(2,2))
# line and point transect examples
# top row points/samplers in space
# bottom row histograms

@@

#par(opar2)
par(mar=c(4, 4, 1, 2) + 0.1)
hist(detected_distances, main="", xlab="Observed distances")

@@

par(mar=c(4, 4, 1, 2) + 0.1)
hist(detected_distances, main="", xlab="Observed distances")

@@

(LT: stranded one-sentence paragraph -- needs integrating.) Figure \ref{fig:pointslines} shows sampling of a population of 500 individuals using line and point transects and their corresponding histograms.

@@

(LT: Only got this far: will do more later...)

@@

We show examples of analyses in \pkg{Distance} using two data sets: one line transect and one point transect. (LT: I think it might be better to delete this next sentence -- after all, two datasets can't possibly represent all the possibilities there are in the real world -- CDS, MCDS, MRDS, with and without covariates, problems such as spiked data, animal movement, measurement error, etc etc.) These data sets have been chosen to be representative of data seen in practice.

@@

The line transect data is based on a survey of Antarctic minke whales (*Balaenoptera bonaerensis*). The data are simulated from models fitted to data from the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua]. They consist of 99 observations and include a stratification based on location (near or distant from ice edge) and effort data (transect lengths). The survey is shown in Figure \ref{fig:minke-strata}.

@@

The two example data sets used here are distributed with \pkg{Distance} so readers can reproduce our results. However, generally data will be collected in the field and will need to be formatted correctly for use with \pkg{Distance}. The package allows for a flexible format for data input ranging from very simple to complex:

>>>>

<!--
(LT: I wonder who the intended audience is for this paper? (LM: - Distance for Windows users and other people looking do distance sampling in R?) The way it is currently written, it ``kinda'' introduces distance sampling as well as the relevant commands.  Is this standard for a paper in this journal -- to introduce the methods as well as the software?  Can we assume a reasonable level of analytical sophistication on the part of the reader, and so start with a few more formulae?  Can we assume they know the methods, more or less, or point them at somewher where this is already introduced.  (Perhaps I should go back and have a look at the EcolApps paper to see what approach we took there... (LM: The paper on Distance in JAppEcol? There is quite a big intro on distance sampling there.))) -->

Distance sampling [@Buckland:2001vm; @Buckland:2004ts; @buckland2015distance] encompasses a suite of methods used to estimate the density and/or abundance of biological populations. Distance sampling can be thought of as an extension of plot sampling. Plot sampling involves selecting a number of plots (small areas) at random within the study area and counting the objects of interest that are contained within each plot. By selecting the plots at random we can assume that the density of objects in the plots is representative of the study area as a whole. One of the key assumptions of plot sampling is that all objects within each of the plots are counted. Distance sampling relaxes this assumption in that observers are no longer required to see and count everything within selected plots. While plot sampling techniques are adequate for static populations occuring at high density they are inefficient for more sparsely distributed populations. Distance sampling provides a more efficient solution in such circumstances.

Conventional distance sampling assumes the observer is either stood at a point or moving along a line and will only see everything that occurs exactly at that point or on the line. It is then expected that the further away an object is from the point or line (also known as the sampler or transect) the less likely it is that the observer will see it. By recording the distance to each of the detected objects we can fit a model to describe the probability of detection given distance from the sampler, which we refer to as the ‘detection function’. The detection function can be used to infer how many objects were missed and thereby produce estimates of density and/or abundance. Further information on distance sampling, including the theory, various extensions to deal with more complex situations and recent methodological developments can be found at http://distancesampling.org/dbib.html

The Windows program Distance (or "DISTANCE"; for clarity henceforth "Distance for Windows") can be used to fit detection functions to such distance sampling data. It was first released in November 1998 as Distance for Windows version 3.5. Since this time it has evolved to include various design and analysis features [@Thomas:2010cf]. Distance for Windows versions 5 onwards have included \proglang{R} [@rcore]  packages as the analysis engines providing additional, more complex analysis options than those offered by the original FORTRAN engine.

As Distance for Windows becomes increasingly reliant on analyses performed in \proglang{R} and new methods are being developed at a rate where it is unfeasible to make all these available in the Windows interface, we are encouraging the use of our \proglang{R} packages directly. In addition, the core packages in \proglang{R} provide a variety of functions useful for data exploration with minimal data formatting. The plotting of a histogram of detection distances only requires these values to be stored in a numerical vector, an exercise that is recommended not only prior to analyses but also early on in the data collection process _(LM need ref here @buckland2015distance ?)_.

Until now those wishing to use our \proglang{R} packages for straight forward distance sampling analyses would have had to negotiate the complex package \pkg{mrds} [@mrds-pkg] designed for mark-recapture distance sampling. In addition, \pkg{mrds} requires a complex data structure before analyses can be completed. \pkg{Distance} is a wrapper package around \pkg{mrds} making it easier to get started with basic distance sampling analyses in \proglang{R}. Like the histogram function, the most basic detection function estimation only requires a numeric vector of distances. Here we demonstrate how to use \pkg{Distance} to fit detection functions, perform model checking and selection, and estimate abundance and/or density.

## Distance sampling in R?

_(LM I wonder if here we should focus on comparing Distance to other R packages offering distance sampling analyses here? Some of the info below could be incorporated into the paragraphs above (I have done this to some extent.. do we need more on distance sampling theory above?)... not done much with this section in case we want to do something different here.)_

<!-- While plot sampling techniques are adequate for static populations occuring at high density they are inefficient for more sparsely distributed populations. Distance sampling provides a more efficient solution in such circumstances. In addition, for cryptic species it simply may not be possible to detect all individuals within a given plot and we must find a way to account for imperfect detectability when estimating abundance [@LahozMonfort:2013bg]. This paper does not provide a detailed explanation of the theory behind distance sampling but instead concentrates on it's application using the package \pkg{Distance}. Further information on distance sampling, including the theory, various extensions to deal with more complex situations and recent methodological developments can be found at http://distancesampling.org/dbib.html -->

Conventional distance sampling analysis simply relies on estimating the probability of detection given how far an object is from the sampler. Note that in the case of populations that occur in groups (clusters) the whole group or cluster is treated as an individual object. Figure \ref{fig:lines} shows example data collection on a simulated population of 500 individuals using line transects and the corresponding histogram of distances recorded by the observer. As expected we see in the histogram of Fig. \ref{fig:lines} that as the perpendicular distance from the transect line increases, the number of detections made in each distance interval decreases. \pkg{Distance} provides a selection of functions to model this probability of detection detailed in Section... below.

```{r lines, fig.height=3, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plot shows an example of a survey of an area containing a population of 500 objects; blue lines indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals. The right side of the figure shows a histogram of observed distances.\\label{fig:lines}"}
opar <- par(mfrow=c(1,2))
# line and point transect examples
# top row points/samplers in space
# bottom row histograms

@@

#par(opar2)
par(mar=c(4, 4, 1, 2) + 0.1)
hist(detected_distances, main="", xlab="Observed perpendicular distances", freq = FALSE)

The probability of detecting an object may not only depend on how far it is from the obsever but also on other factors such as weather conditions, ground cover, cluster size etc. The \pkg{Distance} package also allows the incorporation of such covariates into the modelling process of the detection function.

_(LM How does Distance use the detection function to estimate abundance is it any different to mrds does it infact use dht?)_

<!--

(LM: do we need to talk about assumptions here?)

(LM: detail what Distance deals with?)

Field methods are chosen to be suitable to species and habitat constraints [@buckland2015distance].

```{r points-and-lines, fig.height=6, fig.width=7, echo=FALSE, message=FALSE, fig.cap="Left side plots show an example of a survey of an area containing a population of 500 objects; blue lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals. The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\\label{fig:pointslines}"}
opar <- par(mfrow=c(2,2))
# line and point transect examples
# top row points/samplers in space
# bottom row histograms

#par(opar2)
par(mar=c(4, 4, 1, 2) + 0.1)
hist(detected_distances, main="", xlab="Observed perpendicular distances")

@@

par(mar=c(4, 4, 1, 2) + 0.1)
hist(detected_distances, main="", xlab="Observed radial distances")

@@

(LT: stranded one-sentence paragraph -- needs integrating.)

@@

(LT: Only got this far: will do more later...) -->

@@

We introduce two example analyses performed in \pkg{Distance}: one line transect and one point transect. These data sets have been chosen as they represent typical data seen in practice.

@@

The line transect data have been simulated from models fitted to Antarctic minke whale (*Balaenoptera bonaerensis*) data (_LM why is the Minke whale data simulated rather than ust using original data, just data ownership/permissions?_). These data were collected as part of the International Whaling Commission's International Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer surveys [@Branch:2001ua]. This data set consists of 99 observations that are stratified based on location (near or distant from ice edge) and effort data (transect lengths)(LM did the original dataset have 99 observations or the simulated dataset or both?). The survey is shown in Figure \ref{fig:minke-strata}. _(LM wonder if we need more info on how the data were simulated... based only on detection function or on spatial model too? I notice in the figure text that the simulated data are based on ``1992/93 Area III'' should this info go here too?)_

@@

The two example data sets used here are distributed preformatted with \pkg{Distance} so readers can reproduce our analyses. Generally, data collected in the field will require some formatting for use with \pkg{Distance}. However, the package allows for a flexible format of data input ranging from very simple to complex dependent on the type of analysis you wish to perform:
